Comparative safety and effectiveness of Pfizer BA.4-5 versus Sanofi during the spring 2023 COVID-19 booster vaccination programme in England: a matched cohort study in OpenSAFELY-TPP

### 2024-01-31 write-up/manuscript/manuscript.qmd
update manuscript

---
title: "Comparative effectiveness of Pfizer B.A4-5 versus Sanofi during the Spring 2023 COVID-19 booster programme in England: a cohort study in OpenSAFELY-TPP"
format:
  # html:
  #   toc: false
  #   embed-resources: true
  docx
knitr:
  opts_chunk: 
    collapse: true
    echo: false
    fig.path: ./write-up/figures/
    fig.pos: H
    R.options:
      knitr.graphics.auto_pdf: true
bibliography: references.bib
csl: vancouver.csl
link-citations: true
zotero: yes
header-includes:
- \usepackage{float} 
- \floatplacement{figure}{H}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \captionsetup[table]{labelformat=empty}
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warnings: false
source(here::here("write-up", "manuscript", "setup.R"))
```

```{r}
#| label: process
#| include: false
#| message: false
source(here("write-up", "manuscript", "import-process.R"))
```

William J Hulme^1^, Edward PK Parker^2^, Em Prestige^2^, Ruth H Keogh^2^, Elizabeth J Williamson^2^, Venexia Walker^3,4^, Tom Palmer^3,4^, Helen J Curtis^1^, Alex Walker^1^, Colm Andrews^1^, Amir Mehrkar^1^, Brian MacKenna^1^, Sebastian CJ Bacon^1^, Ben Goldacre^1^, Miguel A Hernán^6,7^\*, Jonathan AC Sterne^3,5,8^\*, and the OpenSAFELY collaborative.

@@

5\. NIHR Bristol Biomedical Research Centre, Bristol, UK

6\. CAUSALab, Harvard T.H. Chan School of Public Health, Boston, MA 02115

7\. Departments of Epidemiology and Biostatistics, Harvard T.H. Chan School of Public Health, Boston, MA 02115

@@

*Introduction* The spring 2023 COVID-19 booster vaccination programme in England used both Pfizer Ba45 and Sanofi vaccines. All people aged 75 years and over, or considered to be clinically vulnerable were eligible to receive a booster dose. Direct comparisons of the effectiveness against severe COVID-19 of these two vaccines for boosting have not been made in trials or observational data.

*Methods* On behalf of NHS England, we used the OpenSAFELY-TPP database to compare recipients of each vaccine, investigating the over 75s and the clinically vulnerable groups separately. Recipients were matched on date of vaccination, vaccine history, age, and other characteristics. Recipients were included if boosted between 1 April and 30 June 2023, and followed up for X weeks. Outcomes were COVId-19 emergency department attendance, COVID-19 hospital admission, and COVID-19 death. We estimated the cumulative incidence of each outcome, and quantified comparative effectiveness using risk differences (RD) and incidence rate ratios (IRRs).

*Results* `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c4") %>% pull(n) %>% sum())` people were 1-1 matched in the CV cohort, and `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart, crit=="c4") %>% pull(n) %>% sum())` in the 75+ cohort, contributing a total `r rst$cv$contrasts_table %>% filter(outcome=="coviddeath", subgroup=="all") %>% pull(persontime_weeks)` and `r rst$age75plus$contrasts_table %>% filter(outcome=="coviddeath", subgroup=="all") %>% pull(persontime_weeks)` person-weeks of follow-up respectively.

In both CV and 75+ cohorts, effectiveness against all measured COVID-19-related outcomes was similar. This finding was observed overall and in pre-specified subgroups.

In the CV cohort, the X-week risks per `r perN_format` of COVID-19 A&E attendance were `r rst$cv$contrasts_table %>% filter(outcome=="covidemergency", subgroup=="all") %>% pull(riskCI_0_95)` for Pfizer B.A4-5 and `r rst$cv$contrasts_table %>% filter(outcome=="covidemergency", subgroup=="all") %>% pull(riskCI_1)` for Sanofi: the IRR comparing Sanofi with PfizerB.A4-5 was `r rst$cv$contrasts_table %>% filter(outcome=="covidemergency", subgroup=="all") %>% pull(irrCI)`. For COVID-19 hospitalisation, X-week risks were `r rst$cv$contrasts_table %>% filter(outcome=="covidadmitted", subgroup=="all") %>% pull(riskCI_0)` and `r rst$cv$contrasts_table %>% filter(outcome=="covidadmitted", subgroup=="all") %>% pull(riskCI_1)`: IRR `r rst$cv$contrasts_table %>% filter(outcome=="covidadmitted", subgroup=="all") %>% pull(irrCI)`. For COVID-19 death: `r rst$cv$contrasts_table %>% filter(outcome=="coviddeath", subgroup=="all") %>% pull(riskCI_0)` and `r contrasts_table %>% filter(outcome=="coviddeath", subgroup=="all") %>% pull(riskCI_1)`: IRR `r rst$cv$contrasts_table %>% filter(outcome=="coviddeath", subgroup=="all") %>% pull(irrCI)`.

In the 75+ cohort, the X-week risks per `r perN_format` for COVID-19 A&E attendance were `r rst$age75plus$contrasts_table %>% filter(outcome=="covidemergency", subgroup=="all") %>% pull(riskCI_0_95)` for Pfizer B.A4-5 and `r rst$age75plus$contrasts_table %>% filter(outcome=="covidemergency", subgroup=="all") %>% pull(riskCI_1)` for Sanofi: the IRR comparing Sanofi with PfizerB.A4-5 was `r rst$age75plus$contrasts_table %>% filter(outcome=="covidemergency", subgroup=="all") %>% pull(irrCI)`. For COVID-19 hospitalisation, X-week risks were `r rst$age75plus$contrasts_table %>% filter(outcome=="covidadmitted", subgroup=="all") %>% pull(riskCI_0)` and `r rst$age75plus$contrasts_table %>% filter(outcome=="covidadmitted", subgroup=="all") %>% pull(riskCI_1)`: IRR `r rst$age75plus$contrasts_table %>% filter(outcome=="covidadmitted", subgroup=="all") %>% pull(irrCI)`. For COVID-19 death: `r perN_format` were `r rst$age75plus$contrasts_table %>% filter(outcome=="coviddeath", subgroup=="all") %>% pull(riskCI_0)` and `r contrasts_table %>% filter(outcome=="coviddeath", subgroup=="all") %>% pull(riskCI_1)`: IRR `r rst$age75plus$contrasts_table %>% filter(outcome=="coviddeath", subgroup=="all") %>% pull(irrCI)`.

*Conclusion* This observational study comparing effectiveness of PFizer and Sanofi vaccine during the Spring 2023 programme in England in the two main eligible cohorts -- people aged 75 and over and in clinically vulnerable people -- found no evidence of substantial differences between the two vaccine products within the first X weeks after vaccination.

@@

```{r break1}
2+2
```

@@

In X 2022, NHS XXX in England announced the Spring 2023 COVID-19 booster vaccination programme, available to those aged 75 years and over, and those considered to be vulnerable to infection or severe COVID-19 disease, based on guidance from the Joint Committee for Vaccination and Immunisation (JCVI) @covid19,. The use of the both the Pfizer-BioNTech B.A.4/5 vaccine and the Sanofi vaccine during the spring 2023 programme enables a direct comparison of their effectiveness against severe COVID-19 disease, which has not yet been conducted in randomised trials.

On behalf of NHS England, we used the OpenSAFELY-TPP database, covering around 45% of English primary care practices and linked to hospital and death records, to compare the effectiveness of boosting with Pfizer and Sanofi in the 75+ years group and the clinically vulnerable group who received either vaccine in the spring 2023 programme. During this period and the weeks following, the X variant was dominant.

@@

All data were linked, stored and analysed securely within the OpenSAFELY platform: <https://opensafely.org/>. With the approval of NHS England, primary care records managed by the GP software provider TPP were linked, using NHS numbers, to A&E attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS). COVID-19 vaccination history and health and social care worker status is available in the GP record directly via the National Immunisation Management System (NIMS).

@@

We considered two cohorts: all adults aged 75 years or over ("75+"), and all adults aged 18 or over who were classified as clinically vulnerable ("CV"). These groups are not mutually exclusive. Clinical vulnerability encompasses various chronic or immunocompromising conditions, immunosuppressive treatments, and ..., and is defined fully in supplementary materials.

In each group, any person receiving a Pfizer or Sanofi vaccine dose between `r study_dates_format$studystart_date` and `r study_dates_format$studyend_date` inclusive were considered for inclusion. People were eligible if they: were registered at a GP practice using TPP's SystmOne clinical information system at the time of boosting; were not a health or social care worker, not resident in a care or nursing home and not medically housebound or receiving end-of-life care; had no evidence of SARS-CoV-2 infection or COVID-19 disease within the previous 90 days; were not hospitalised at the time of boosting; and had complete information on sex, deprivation, and Sustainability and Transformation Partnership (STP, a geographical grouping of NHS and local authorities).

@@

Pfizer and Sanofi booster recipients were matched 1-1 without replacement, on the following characteristics: date of vaccination; number of previous COVID-19 vaccine doses; time since previous COVID-19 vaccine dose; sex (male or female); age (3 year caliper and within age groups defined by JVCI risk groups); Index of Multiple Deprivation (IMD, grouped by quintile); STP as a surrogate for geographical region; evidence of prior SARS-CoV-2 infection (any of positive SARS-CoV-2 test, probable infection documented in primary care, or COVID-19 hospital attendance or admission); and morbidity count (grouped as 0, 1, or 2 or more conditions from the following list: diabetes, BMI over 40kg/m^2^, chronic heart disease, chronic kidney disease, chronic liver disease, chronic respiratory disease or severe asthma, chronic neurological disease, cancer within 3 years). Clinical vulnerability was an additional matchng factor in the 75+ cohort.

There are no missing values for these variables: people with missing values for sex, IMD or STP were excluded, and all other variables were defined by the presence or absence of clinical codes or events.The supplementary materials provide more information on how these characteristics were chosen and defined.

@@

We considered three primary outcomes. COVID-19 emergency department admission was identified using HES emergency department records, with SNOMIED.

COVID-19 hospitalisation was identified using Secondary Use Service (SUS) in-patient hospital records with International Statistical Classification of Diseases and Related Health Problems 10th Revision (ICD-10) codes U07.1, U07.2, or U.109 as the primary or non-primary reason for admission. COVID-19 death was defined as death with U07.1, U07.2, or U.109 ICD-10 codes mentioned anywhere on the death certificate (i.e., as an underlying or contributing cause of death). We also report on non-COVID-19 deaths.

<!-- We also considered the follows safety outcomes: -->

<!-- * Adverse allergic reaction / Anaphylaxis -->

<!-- * Myocarditis or Pericarditis -->

<!-- *  -->

<!--# COVID-19 related A&E attendance and critical/intensive care admission we also considered but there were doubts over data the reliability of event ascertainment in these cases. -->

Each person was followed from receipt of vaccination (time zero) until the outcome of interest, with censoring at death, practice de-registration, XX weeks, or `r study_dates_format$followupend_date`.

@@

Baseline characteristics of each vaccine group were tabulated, with between-group balance examined using standardised mean differences. We estimated the cumulative incidence of each outcome in each vaccine group using the Kaplan-Meier (KM) estimator. We estimated cumulative risk differences and risk ratios (RRs) comparing the vaccine groups for each outcome. These were estimated up to 20 weeks where available, but estimates at X weeks are reported as the primary comparison to ensure potential follow-up in those vaccinated towards the end of the booster programme was not censored. We also estimated Incidence Rate Ratios (IRRs) over the first X weeks, and within period-specific intervals defined by splits on weeks 1, 2, 4, 8, 12, 16, and X, where available.

<!-- days 7, 14, 21, 28, 42, 56, 70, 84, 112, 140, 168 and 196. -->

@@

95% confidence limits for the cumulative incidences were derived using standard errors on the log-KM scale. Confidence limits for the risk differences were derived from the sum of squares of the standard errors on the KM scale, using Greenwood's formula. Confidence limits for the risk ratio were derived from the sum of squares of the standard errors on the log-KM scale. Confidence limits for the IRRs were derived from the sum of squares of the standard errors on the log-KM scale. <!-- 95% confidence limits for the Kaplan-Meier estimates, risk differences, risk ratios, and incidence rate ratios are derived by bootstrapping the matched pairs. 500 bootstrap replicates will be used. -->

@@

We estimated comparative effectiveness separately in the following subgroups: age band in the CV group only (18-39, 40-54, 55-64, 65-74, 75+); clinical vulnerability in the 75+ group only; prior vaccination history; evidence of prior SARS-CoV-2 infection or not. We used $\chi^2$ tests to examine evidence for heterogeneity in comparative effectiveness between subgroups.

We reported the rates of SARS-CoV-2 tests undertaken from baseline up to 31 March 2022 in each vaccine group, as lower testing rates in one group might lead to under-ascertainment of outcomes, particularly positive SARS-CoV-2 tests.

We reported outcomes at one week as a negative control outcome @Lipsitch2010, during which time the vaccine is not expected to have any protective immunological effect.

@@

```{r break2}
2+2
```

@@

\[\[fix total identified in data_process script then re-write\]\]

<!-- We identified `r label_number(1, big.mark=",")(filter(rst$cv$total_flowchart, boost_type=="Any") %>% pull(n) %>% sum)` clinically vulnerable people and `r label_number(1, big.mark=",")(filter(rst$age75plus$total_flowchart, boost_type=="Any") %>% pull(n) %>% sum)` people aged 75 or over who were recorded as having received a COVID-19 vaccine between `r study_dates_format$studystart_date` and `r study_dates_format$studyend_date` whilst registered at a TPP practice. This included `r label_number(1, big.mark=",")(filter(rst$age75plus$total_flowchart, boost_type=="pfizerBA45") %>% pull(n))` (`r label_percent(0.1)(filter(rst$cv$total_flowchart, boost_type=="pfizerBA45") %>% pull(pct))`) Pfizer B.A4/5 and `r label_number(1, big.mark=",")(filter(rst$cv$total_flowchart, boost_type=="sanofi") %>% pull(n))` (`r label_percent(0.1)(filter(rst$cv$total_flowchart, boost_type=="sanofi") %>% pull(pct))`) Sanofi. -->

<!-- `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c0") %>% pull(n) %>% sum())` were in the CV cohort, and of these `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c1", boost_type=="pfizerBA45") %>% pull(n))` (`r label_percent(0.1)(filter(rst$cv$flowchart, crit=="c1", boost_type=="pfizerBA45") %>% pull(pct_all))`) received Pfizer BA45 and `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c1", boost_type=="sanofi") %>% pull(n))` (`r label_percent(0.1)(filter(rst$cv$flowchart, crit=="c1", boost_type=="sanofi") %>% pull(pct_all))`) received Sanofi, with `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c3", boost_type=="pfizerBA45") %>% pull(n))` (`r label_percent(0.1)(filter(rst$cv$flowchart, crit=="c3", boost_type=="pfizerBA45") %>% pull(pct_all))`) and `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c3", boost_type=="sanofi") %>% pull(n))` (`r label_percent(0.1)(filter(rst$cv$flowchart, crit=="c3", boost_type=="sanofi") %>% pull(pct_all))`) eligible for matching respectively (Figure 1). -->

`r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart, crit=="c0") %>% pull(n) %>% sum())` were in the 75+ cohort, and of these `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart, crit=="c1", boost_type=="pfizerBA45") %>% pull(n))` (`r label_percent(0.1)(filter(rst$age75plus$flowchart, crit=="c1", boost_type=="pfizerBA45") %>% pull(pct_all))`) received Pfizer BA45 and `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart, crit=="c1", boost_type=="sanofi") %>% pull(n))` (`r label_percent(0.1)(filter(rst$age75plus$flowchart, crit=="c1", boost_type=="sanofi") %>% pull(pct_all))`) received Sanofi, with `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart, crit=="c3", boost_type=="pfizerBA45") %>% pull(n))` (`r label_percent(0.1)(filter(rst$age75plus$flowchart, crit=="c3", boost_type=="pfizerBA45") %>% pull(pct_all))`) and `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart, crit=="c3", boost_type=="sanofi") %>% pull(n))` (`r label_percent(0.1)(filter(rst$age75plus$flowchart, crit=="c3", boost_type=="sanofi") %>% pull(pct_all))`) eligible for matching respectively (Figure 1).

Eligible Pfizer BA45 recipients were on average older, more deprived and had higher rates of prior clinical conditions than Sanofi recipients (supplementary Table S1). They were also more likely to have received more/less/X prior COVID-19 vaccine doses...

In the CV cohort, `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c4", boost_type=="pfizerBA45") %>% pull(n))` people in each vaccine group (total `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c4", boost_type %in% c("pfizerBA45", "sanofi")) %>% pull(n) %>% sum())`) were matched, representing `r label_percent(0.1)(filter(rst$cv$flowchart, crit=="c4", boost_type=="pfizerBA45") %>% pull(pct_step))` and `r label_percent(0.1)(filter(rst$cv$flowchart, crit=="c4", boost_type=="sanofi") %>% pull(pct_step))` of eligible Pfizer and Sanofi recipients respectively (Figure 1, Figure S1).

>>>>

---
title: "Comparative safety and effectiveness of Pfizer BA.4-5 versus Sanofi during the spring 2023 COVID-19 booster vaccination programme in England: a matched cohort study in OpenSAFELY-TPP"
format:
  # html:
  #   toc: false
  #   embed-resources: true
  rtf
output-file: manuscript_draft.rtf
keep-md: false
execute:
  echo: false
  output: true
  #warning: true
knitr:
  opts_chunk: 
    #fig.path: ./
    fig.pos: H
    R.options:
      knitr.graphics.auto_pdf: true
bibliography: references.bib
csl: vancouver.csl
link-citations: true
zotero: true
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warnings: false
#| cache: false
source(here::here("write-up", "manuscript", "setup.R"))
```

```{r}
#| label: process
#| include: false
#| message: false
#| cache: false
source(here("write-up", "manuscript", "import-process.R"))
```

TBD: Colm Andrews^1^, Edward PK Parker^2^, Em Prestige^2^, Venexia Walker^3,4^, Tom Palmer^3,4^, Helen J Curtis^1^, Alex Walker^1^, Amir Mehrkar^1^, Brian MacKenna^1^, Sebastian CJ Bacon^1^, Ben Goldacre^1^, Miguel A Hernán^5,6^\*, Jonathan AC Sterne^3,7,8^\*, William J Hulme^1^, and the OpenSAFELY collaborative.

@@

5\. CAUSALab, Harvard T.H. Chan School of Public Health, Boston, MA 02115

6\. Departments of Epidemiology and Biostatistics, Harvard T.H. Chan School of Public Health, Boston, MA 02115

7\. NIHR Bristol Biomedical Research Centre, Bristol, UK

@@

*Introduction* The spring 2023 COVID-19 booster vaccination programme in England used both Pfizer BA.4-5 and Sanofi vaccines. All people aged 75 years or over and the clinically vulnerable were eligible to receive a booster dose. Direct comparisons of effectiveness against severe COVID-19 events of these two vaccines for boosting have not been made in trials or observational data.

*Methods* With the approval of NHS England, we used the OpenSAFELY-TPP database to compare effectiveness of the Pfizer BA.4-5 and Sanofi vaccines during the spring 2023 booster programme, between 1 April and 30 June 2023. We investigated two cohorts separately: those aged 75 or over (75+); those aged 50 or over and clinically vulnerable (CV). In each cohort, vaccine recipients were matched on date of vaccination, COVID-19 vaccine history, age, and other characteristics. Effectiveness outcomes were COVID-19 Emergency Department (ED) attendance, COVID-19 hospital admission, and COVID-19 death up to 16 weeks after vaccination. Safety outcomes were Pericarditis and Myocarditis up to 4 weeks after vaccination. We report the cumulative incidence of each outcome, and compare safety and effectiveness using risk differences (RD), relative risks (RR), and incidence rate ratios (IRRs).

*Results* `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c4") %>% pull(n) %>% sum())` people were 1-1 matched in the CV cohort, and `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart, crit=="c4") %>% pull(n) %>% sum())` in the 75+ cohort, contributing a total `r rst$cv$contrasts_table %>% filter(outcome=="coviddeath", subgroup=="all") %>% pull(persontime_weeks)` and `r rst$age75plus$contrasts_table %>% filter(outcome=="coviddeath", subgroup=="all") %>% pull(persontime_weeks)` person-weeks of follow-up respectively.
The incidence of COVID-19 hospital admission was higher for Sanofi than for Pfizer BA.4-5. In the CV cohort, 16-week risks per `r perN_format` people were `r rst$cv$contrasts_table %>% filter(outcome=="covidadmitted", subgroup=="all") %>% pull(riskCI_0_95)` for Pfizer BA.4-5 and `r rst$cv$contrasts_table %>% filter(outcome=="covidcritcare", subgroup=="all") %>% pull(riskCI_1)` for Sanofi, with an IRR of `r rst$cv$contrasts_table %>% filter(outcome=="covidadmitted", subgroup=="all") %>% pull(irrCI_95)`. In the 75+ cohort, these were `r rst$cv$contrasts_table %>% filter(outcome=="covidadmitted", subgroup=="all") %>% pull(riskCI_0)` for Pfizer BA.4-5 and `r rst$cv$contrasts_table %>% filter(outcome=="covidadmitted", subgroup=="all") %>% pull(riskCI_1)` for Sanofi, with an IRR of `r rst$cv$contrasts_table %>% filter(outcome=="covidadmitted", subgroup=="all") %>% pull(irrCI)`. These finding were similar across all pre-specified subgroups.
More severe COVID-19 related outcomes (critical care admission and death), and safety outcomes at 4 weeks, were rare in both vaccines so we could not reliably estimate superiority of either vaccine.

*Conclusion* This observational study comparing effectiveness of Pfizer BA.4-5 and Sanofi vaccine during the spring 2023 programme in England in the two main eligible cohorts -- people aged 75 and over and in clinically vulnerable people -- found some evidence of superior effectiveness against COVID-19 hospital admission for Pfizer BA.4-5 compared with Sanofi within 16 weeks after vaccination. Differences were largely observed during weeks 2 to 4 after vaccination, suggesting the Pfizer BA.4-5 offers earlier protection via more rapid immunogenicity, as shown in RCTs.

@@

````{comment}

````

@@

The spring 2023 COVID-19 booster vaccination programme in England offered COVID-19 vaccination to those aged 75 years or over, resident in a care home, and those who are immunosuppressed, based on guidance from the Joint Committee for Vaccination and Immunisation (JCVI).[@NHSEnglandNHSc; @JCVIStatementCOVID19; @JCVIStatementSpring] Both the Pfizer BA.4/BA.5 vaccine and the Sanofi vaccine were used during this programme. This allows direct comparisons of their safety and effectiveness against severe COVID-19 disease, something not investigated in randomised trials.

We used the OpenSAFELY-TPP database, covering around 45% of English primary care practices and linked to hospital and death records, to compare safety and effectiveness of the Pfizer BA.4-5 and Sanofi vaccines received during the spring 2023 booster programme.

@@

All data were linked, stored and analysed securely using the OpenSAFELY platform: <https://opensafely.org/>, as part of the NHS England OpenSAFELY COVID-19 service. Primary care records managed by the GP software provider, TPP were linked, using NHS numbers, to emergency department attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS). COVID-19 vaccination history and health and social care worker status is available in the GP record directly via the National Immunisation Management System (NIMS).

@@

We considered two cohorts: all adults aged 75 years or over ("75+"), and all adults aged 50 or over who were considered clinically vulnerable ("CV"). These groups are not mutually exclusive. We only considered those aged 50 years or over because the Sanofi vaccine was rarely offered to younger people. Clinical vulnerability encompasses various chronic or immunosuppressive conditions and medications as set out in the Green Book chapter 14a @COVID19GreenBook, and defined in supplementary materials.

In each group, any person receiving a Pfizer BA.4-5 or Sanofi vaccine dose between `r study_dates_format$studystart_date` and `r study_dates_format$studyend_date` inclusive was considered. People were eligible if they: were registered at a GP practice using TPP's SystmOne clinical information system at the time of vaccination; had received at least two prior COVID-19 vaccine doses; were not a health or social care worker and not receiving end-of-life care; had no evidence of SARS-CoV-2 infection or COVID-19 disease during the 28 days prior to vaccination; were not a hospital in-patient at the time of vaccination; and had complete information on sex, deprivation, and Sustainability and Transformation Partnership (STP, a geographical grouping of NHS and Local Authorities).

@@

Pfizer BA.4-5 and Sanofi vaccine recipients were matched 1-1 without replacement, on the following characteristics: date of vaccination (3-day caliper); number of previous COVID-19 vaccine doses (grouped as 2 to 4, 5, or 6 or more); time since previous COVID-19 vaccine dose (28-day caliper); sex (male or female); age (3-year caliper and within age groups 50-64, 65-75, 75-79, 80-84, 85+); Index of Multiple Deprivation (IMD, within a 5000 IMD-rank caliper); STP as a surrogate for geographical region; any evidence of prior SARS-CoV-2 infection (including any of: positive SARS-CoV-2 test, infection as documented in primary care, COVID-19 hospital attendance or admission); and morbidity count (grouped as 0, 1, or 2 or more conditions from the following list: diabetes, BMI over 40kg/m^2^, chronic heart disease, chronic kidney disease, chronic liver disease, chronic respiratory disease or severe asthma, chronic neurological disease, cancer within 3 years). Clinical vulnerability was an additional matching factor in the 75+ cohort.

Age was calculated as at 30 June 2023, to allow for a degree of operational flexibility in who was offered vaccination during the programme. All other characteristics were calculated as on the day before vaccination. In the analysis cohorts there are no missing values for any characteristics as people with missing values for sex, IMD or STP were excluded, and remaining characteristics were defined by the presence or absence of clinical codes or events in the health record. The supplementary materials provide more information on how these characteristics were defined.

@@

We considered three effectiveness outcomes. COVID-19 hospital admission was identified using Secondary Use Service (SUS) in-patient hospital spell records with International Statistical Classification of Diseases and Related Health Problems 10th Revision (ICD-10) codes U07.1, U07.2, or U.109 as the primary or non-primary reason for admission. COVID-19 critical care admissions were any COVID-19 hospital spell with at least one day spent in ciritcal care (including any such spells occurring subsequent to the first spell after vaccination). COVID-19 death was defined as death COVID-19 ICD-10 codes (as before) mentioned anywhere on the death certificate (i.e., as an underlying or contributing cause of death). We also report on non-COVID-19 deaths.

We considered two safety outcomes. Pericarditis within 28 days of vaccination was identified using SUS emergency department records with pericarditis listed as the discharge diagnosis (SNOMED), and using ICD-10 coded hospital admission and death records. Myocarditis was identified similarly. Details of the codelists used to define these outcomes are available in supplementary materials.

<!--# COVID-19 related attendance and critical/intensive care admission we also considered but there were doubts over data the reliability of event ascertainment in these cases. -->

Each person was followed from receipt of vaccination (time zero) until the outcome of interest, with censoring at death, practice de-registration, `r study_dates_format$followupend_date`. Outcomes were observed for up to 16 weeks for effectiveness outcomes, and up to 4 weeks for safety outcomes.

@@

Baseline characteristics of each vaccine group were tabulated, with between-group balance examined using standardised mean differences. We estimated the cumulative incidence of each outcome in each vaccine group using the Kaplan-Meier (KM) estimator. We estimated cumulative risk differences and risk ratios (RRs) comparing the vaccine groups for each outcome. We also estimated Incidence Rate Ratios (IRRs) over the full follow-up duration, and within period-specific intervals defined by splits on weeks 1, 2, 4, 8, 12, and 16.

@@

95% confidence limits for the cumulative incidences were derived using standard errors on the log-KM scale. Confidence limits for the risk differences were derived from the sum of squares of the standard errors on the KM scale, using Greenwood's formula. Confidence limits for the risk ratios and incidence rate ratios were derived from the sum of squares of the standard errors on the log-KM scale. <!-- 95% confidence limits for the Kaplan-Meier estimates, risk differences, risk ratios, and incidence rate ratios are derived by bootstrapping the matched pairs. 500 bootstrap replicates will be used. -->

@@

We estimated comparative effectiveness separately in the following subgroups: age band in the CV group only (50-64, 65-75, 75-79, 80-84, 85+); clinical vulnerability in the 75+ group only; prior COVID-19 vaccine count. We used $\chi^2$ tests to examine evidence for heterogeneity in estimates between subgroups.

<!-- We reported outcomes at one week as a negative control outcome @Lipsitch2010, during which time the vaccine is not expected to have any protective immunological effect. -->

@@

````{comment}

````

@@

We identified `r label_number(1, big.mark=",")(filter(flowchart_totals_allcohorts, boost_type=="any") %>% pull(n))` people aged 50 years or over who were recorded as having received a COVID-19 vaccine between `r study_dates_format$studystart_date` and `r study_dates_format$studyend_date` whilst registered at a TPP practice. This included `r label_number(1, big.mark=",")(filter(flowchart_totals_allcohorts, boost_type=="pfizerBA45") %>% pull(n))` (`r label_percent(0.1)(filter(flowchart_totals_allcohorts, boost_type=="pfizerBA45") %>% pull(pct))`) Pfizer BA.4-5 and `r label_number(1, big.mark=",")(filter(flowchart_totals_allcohorts, boost_type=="sanofi") %>% pull(n))` (`r label_percent(0.1)(filter(flowchart_totals_allcohorts, boost_type=="sanofi") %>% pull(pct))`) Sanofi recipients.

We identified `r label_number(1, big.mark=",")(filter(rst$cv$flowchart_totals, boost_type=="any") %>% pull(n) %>% sum())` people in the CV cohort, and of these `r label_number(1, big.mark=",")(filter(rst$cv$flowchart_totals, boost_type=="pfizerBA45") %>% pull(n))` (`r label_percent(0.1)(filter(rst$cv$flowchart_totals, boost_type=="pfizerBA45") %>% pull(pct))`) received Pfizer BA.4-5 and `r label_number(1, big.mark=",")(filter(rst$cv$flowchart_totals, boost_type=="sanofi") %>% pull(n))` (`r label_percent(0.1)(filter(rst$cv$flowchart_totals, boost_type=="sanofi") %>% pull(pct))`) received Sanofi, with `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c3", boost_type=="pfizerBA45") %>% pull(n))` (`r label_percent(0.1)(filter(rst$cv$flowchart, crit=="c3", boost_type=="pfizerBA45") %>% pull(pct_all))`) and `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c3", boost_type=="sanofi") %>% pull(n))` (`r label_percent(0.1)(filter(rst$cv$flowchart, crit=="c3", boost_type=="sanofi") %>% pull(pct_all))`) eligible for matching respectively (Figure 1).

`r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart_totals, boost_type=="any") %>% pull(n) %>% sum())` people were in the 75+ cohort, and of these `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart_totals, boost_type=="pfizerBA45") %>% pull(n))` (`r label_percent(0.1)(filter(rst$age75plus$flowchart_totals, boost_type=="pfizerBA45") %>% pull(pct))`) received Pfizer BA.4-5 and `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart_totals, boost_type=="sanofi") %>% pull(n))` (`r label_percent(0.1)(filter(rst$age75plus$flowchart_totals, boost_type=="sanofi") %>% pull(pct))`) received Sanofi, with `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart, crit=="c3", boost_type=="pfizerBA45") %>% pull(n))` (`r label_percent(0.1)(filter(rst$age75plus$flowchart, crit=="c3", boost_type=="pfizerBA45") %>% pull(pct_all))`) and `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart, crit=="c3", boost_type=="sanofi") %>% pull(n))` (`r label_percent(0.1)(filter(rst$age75plus$flowchart, crit=="c3", boost_type=="sanofi") %>% pull(pct_all))`) eligible for matching respectively (Figure 1).

Sanofi was more likely to be administered in April 2023, and Pfizer BA.4-5 more likely in May and June 2023.
Prior to matching, Sanofi recipients were on average older, more likely to be recorded as resident in a care or nursing home, and less likely to have received fewer than 5 prior COVID-19 vaccine doses compared with Pfizer BA.4-5 recipients. Sanofi recipients were less likely that Pfizer BA.4-5 recipients to have previously received the Pfizer/BA.1 vaccine during an earlier booster programme and more likely to have received Moderna/Omicron. Sex, deprivation levels, and ethnicity were similar in both groups prior to matching (supplementary Tables S1a and S1b).

In the CV cohort, `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c4", boost_type=="pfizerBA45") %>% pull(n))` people in each vaccine group (total `r label_number(1, big.mark=",")(filter(rst$cv$flowchart, crit=="c4", boost_type %in% c("pfizerBA45", "sanofi")) %>% pull(n) %>% sum())`) were matched, representing `r label_percent(0.1)(filter(rst$cv$flowchart, crit=="c4", boost_type=="pfizerBA45") %>% pull(pct_step))` and `r label_percent(0.1)(filter(rst$cv$flowchart, crit=="c4", boost_type=="sanofi") %>% pull(pct_step))` of eligible Pfizer and Sanofi recipients respectively (Figure 1, Figure S1a).
In the 75+ cohort, `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart, crit=="c4", boost_type=="pfizerBA45") %>% pull(n))` were matched in each vaccine group (total `r label_number(1, big.mark=",")(filter(rst$age75plus$flowchart, crit=="c4", boost_type %in% c("pfizerBA45", "sanofi")) %>% pull(n) %>% sum())`), representing `r label_percent(0.1)(filter(rst$age75plus$flowchart, crit=="c4", boost_type=="pfizerBA45") %>% pull(pct_step))` and `r label_percent(0.1)(filter(rst$age75plus$flowchart, crit=="c4", boost_type=="sanofi") %>% pull(pct_step))` of eligible Pfizer and Sanofi recipients respectively (Figure 1, Figure S1b).

After matching, characteristics at the start of follow-up were well-balanced between groups (Table 1) with standardised mean differences consistently under 0.05 (supplementary Figures S2a and S2b), including for those characteristics not used for matching.
